Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-26 paper/main.tex
add and fix citations

We develop a new agent-based computational model for flexible, elongated bacteria which
is a mechanical feature that is absent in most \acp{abm}.
So far, only a few computational studies have considered flexible rods~\cite{Duman2018,Joshi2019},
but their models cannot be transferred directly to more complex biological systems which might
involve intracellular reactions, extracellular compounds, growth and division.
Only a few \acp{abm} support even the less complex cylindrical
shapes~\cite{Kang2014,breitwieser_biodynamo_2022,Gorochowski2012} but none are able to capture
flexible filamentous bacteria.
This poses a problem in the computational study of these systems.

@@

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
We follow a phenomenological approach thus aiming to fit our model to given datapoints and determine
the type of interaction in this way.
In order to accurately depict the physical size of the rod, our interactions will be repulsive for
short ranges $r<R_1+R_2$ and attractive at longer ranges.
The Mie potential~\cite{Mie1903} is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels corresponding to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower number of parameters and more restrictive mathematical formulation
but may therefore be advantagous in situations of sparse data.
These potentials are a good choice for a simple, interpretable interaction potential with
low number of parameters that can be estimated.

>>>>

We develop a new agent-based computational model for flexible, elongated bacteria which
is a mechanical feature that is absent in most \acp{abm}.
So far, only a few computational studies have considered flexible rods~\cite{Duman2018,Joshi2019},
but their models cannot be transferred directly to more complex biological systems which might
involve intracellular reactions, extracellular compounds, growth and division.
Only a few \acp{abm} support even the less complex cylindrical
shapes~\cite{Kang2014,Breitwieser2021,Gorochowski2012} but none are able to capture
flexible filamentous bacteria.
This poses a problem in the computational study of these systems.

@@

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
We follow a phenomenological approach thus aiming to fit our model to given datapoints and determine
the type of interaction in this way.
In order to accurately depict the physical size of the rod, our interactions will be repulsive for
short ranges $r<R_1+R_2$ and attractive at longer ranges.
The Mie potential~\cite{Mie1903} is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels corresponding to overlaps.
In contrast, an alternative which has been used by other \acp{abm} is the Morse
potential~\cite{Morse1929,Kang2014,Breitwieser2021}, which is less
customizable due to the lower number of parameters and more restrictive mathematical formulation
but may therefore be advantagous in situations of sparse data.
These potentials are a good choice for a simple, interpretable interaction potential with
low number of parameters that can be estimated.
